Why does preregistration increase the persuasiveness of evidence? A Bayesian rationalization.

### 2022-08-19 manuscript.Rmd
Update manuscript.Rmd

We already have noted that there is not much to do about prior probability ($P(H)$, and hence its counter probability $P(¬H)$), and that it is common sense to increase detectability ($P(E|H)$).
The real lever to pull is, therefore, $P(E|¬H)$.
This probability tells us how likely it is that we find evidence in favor of the theory when in fact, the theory is not true.
Its counter probability $P(¬E|¬H)= 1 - P(E|¬H)$ is what we call "theoretical risk", because it is the risk a theory takes on in predicting the occurrence of particular evidence in its favour.
We "borrow" the term from @meehlTheoreticalRisksTabular1978, though he has not assigned it to the probability $P(¬E|¬H)$.
In a response to the original paper, @kuklaClinicalStatisticalTheory1990 argued that the arguments layed out in @meehlTheoreticalRisksTabular1978 can be reconstructed in purely Bayesian framework.
However, while he did not name $P(¬E|¬H)$ but did suggest that @meehlTheoreticalRisksTabular1978 used the term "very strange coincidence" for a small $P(E|¬H)$ which would imply that $P(¬E|¬H)$ can be related to or even equated to theoretical risk.

Let us note some interesting properties of theoretical risk ($P(¬E|¬H)$).
First, increasing theoretical risk leads to higher posterior probability ($P(H|E)$, our objective).
Second, if the theoretical risk is smaller than detectability ($P(E|H)$) it follows that the posterior probability must decrease when observing the evidence.
If detectability exceeds theoretical risk, the evidence is less likely under the theory than it is when the theory does not hold.
Third, if the theoretical risk equals zero, then posterior probability is at best equal to prior probability but only if detectability is perfect ($P(H|E)$ = 1).
In other words, observing a sure fact does not lend credence to a hypothesis.

@@

Both theoretical risk ($P(¬E|¬H)$) and detectability ($P(E|H)$) aggregate uncountable influences, otherwise they could not model the process of evidential support for theories.
To illustrate the concepts we introduced up to here, consider the following  example of a single theory and three experiments that may test it. 
The experiments were created to illustrate how they may differ in their theoretical risk and detectability.
Suppose the primary theory is about the cognitive phenomenon of "insight."
For the purpose of illustration, we define it, with quite some hand-waving, as a cognitive abstraction that allows agents to consistently solve a well-defined class of problems.
We present the hypothesis that the following problem belongs to such a class of insight problems:

@@

We argue that experiment 1 has high theoretical risk ($P(¬E_1|¬H)$) and high detectability ($P(E_1|H)$).
If "insight" has nothing to do with solving the problem ($¬H$), then presenting the insight that Roman numerals could be used, should not lead to all students solving the problem ($¬E_1$); the experiment therefore has high theoretical risk ($P(¬E_1|¬H)$)
Conversely, if insight is required to solve the problem ($H$), then it is likely to help all students to solve the problem ($E_1$), the experiment therefore has high detectability ($P(E_1|H)$).
The second experiment, on the other hand, has low theoretical risk ($P(¬E_2|¬H)$).
Even if "insight" has nothing to do with solving the problem ($¬H$), there are other plausible reasons for observing the evidence ($E_2$), because the students could simply copy the solution, without having any insight.
With regard to detectability, experiments 1 and 2 differ in no obvious way.
Experiment 3, however, also has low detectability.
It is unlikely that all students come up with the correct solution in a short time ($E_3$), even if insight is required ($H$) experiment 3 therefore has low detectability ($P(E_3|H)$).
The theoretical risk, however, is also low in absolute terms, but high compared to the detectability.
In the unlikely event that all 10 students place their matches to form the Roman numeral Ⅷ ($E_3$), it is probably due to insight ($H$) and not by chance $P(¬E_2|¬H)$).
Of course, in practice, we would allow the evidence to be probabilistic, e.g., relax the requirement of "all students" to nine out of ten students, more than eight, and so forth.

As argued earlier, the remainder of the paper will focus on binary, non-probabilisitic evidence to keep the mathematical notation as simple as possible.
We discuss the relation between statistical methods and theoretical risk in the [Statistical Methods] section.

>>>>

We already have noted that there is not much to do about prior probability ($P(H)$ and that it is common sense to increase detectability ($P(E|H)$). <!-- the question is not whether it is common sense, but whether it has something to do with preregistration -->
The real lever to pull is, therefore, $P(E|¬H)$.
This probability tells us how likely it is that we find evidence in favor of the theory when in fact, the theory is not true.
Its counter probability $P(¬E|¬H)= 1 - P(E|¬H)$ is what we call "theoretical risk", because it is the risk a theory takes on in predicting the occurrence of particular evidence in its favour. <!-- I am not shure if thats why we call it theoretical risk. I am also not shure if this is gramatically sensible. -->
We "borrow" the term from @meehlTheoreticalRisksTabular1978, though he has not assigned it to the probability $P(¬E|¬H)$.
In a response to the original paper, @kuklaClinicalStatisticalTheory1990 argued that the arguments layed out in @meehlTheoreticalRisksTabular1978 can be reconstructed in purely Bayesian framework.
However, while he did not name $P(¬E|¬H)$ but did suggest that @meehlTheoreticalRisksTabular1978 used the term "very strange coincidence" for a small $P(E|¬H)$ which would imply that $P(¬E|¬H)$ can be related to or even equated to theoretical risk.

Let us note some interesting properties of theoretical risk ($P(¬E|¬H)$).
First, increasing theoretical risk leads to higher posterior probability ($P(H|E)$, our objective).
Second, if the theoretical risk is smaller than the detectability ($P(E|H)$) it follows that the posterior probability must decrease when observing the evidence.
If detectability exceeds theoretical risk, the evidence is less likely under the theory than it is when the theory does not hold.
Third, if the theoretical risk equals zero, then posterior probability is at best equal to prior probability but only if detectability is perfect ($P(H|E)$ = 1).
In other words, observing a fact that would also be observed if the hypotheses was wrong does not lend credence to the hypothesis.

@@

Both theoretical risk ($P(¬E|¬H)$) and detectability ($P(E|H)$) aggregate uncountable influences, otherwise they could not model the process of evidential support for theories. <!-- I dont understand this sentence -->
To illustrate the concepts we introduced up to here, consider the following  example of a single theory and three experiments that may test it.
These examples were created to illustrate how experiments may differ in their theoretical risk and detectability.
Suppose the primary theory is about the cognitive phenomenon of "insight."
For the purpose of illustration, we define it, with quite some hand-waving, as a cognitive abstraction that allows agents to consistently solve a well-defined class of problems.
We present the hypothesis that the following problem belongs to such a class of insight problems:

@@

We argue that experiment 1 has high theoretical risk ($P(¬E_1|¬H)$) and high detectability ($P(E_1|H)$).
If "insight" has nothing to do with solving the problem ($¬H$), then presenting the insight that Roman numerals could be used, should not lead to all students solving the problem ($¬E_1$); the experiment therefore has high theoretical risk ($P(¬E_1|¬H)$)
Conversely, if insight is required to solve the problem ($H$), then it is likely to help all students to solve the problem ($E_1$), the experiment therefore has high detectability ($P(E_1|H)$).
The second experiment, on the other hand, has low theoretical risk ($P(¬E_2|¬H)$).
Even if "insight" has nothing to do with solving the problem ($¬H$), there are other plausible reasons for observing the evidence ($E_2$), because the students could simply copy the solution, without having any insight.
With regard to detectability, experiments 1 and 2 differ in no obvious way.
Experiment 3, however, also has low detectability.
It is unlikely that all students come up with the correct solution in a short time ($E_3$), even if insight is required ($H$) experiment 3 therefore has low detectability ($P(E_3|H)$).
The theoretical risk, however, is also low in absolute terms, but high compared to the detectability.
In the unlikely event that all 10 students place their matches to form the Roman numeral Ⅷ ($E_3$), it is probably due to insight ($H$) and not by chance $P(¬E_2|¬H)$).
Of course, in practice, we would allow the evidence to be probabilistic, e.g., relax the requirement of "all students" to nine out of ten students, more than eight, and so forth.
<!-- I think this example needs to be swapped for a better one. For example: H: A coin is fair. E1: Observe if the person throwing the coin is betwenn 0 and 10 meters in height (high detectability, no theoretical risk). E2: Throw the coin 2 times and observe if it lands one time on heads and one time on
As argued earlier, the remainder of the paper will focus on binary, non-probabilisitic evidence to keep the mathematical notation as simple as possible.
We discuss the relation between statistical methods and theoretical risk in the [Statistical Methods] section.
